---
title: "Explore GSE76275"
output: 
  html_notebook:
    toc: true
    toc_depth: 2
    toc_float: true
---

# Loading libraries

```{r}
library(GEOquery)
library(oligo)
library(sva)
library(tidyverse)
library(ggsci)
library(factoextra)
library(pheatmap)
library(caret)
library(RColorBrewer)
library(viridis)
library(UpSetR)
library(ComplexUpset)
```


# Custom functions

```{r}
# given a matrix, perform min-max scaling on its columns
min_max_mat <- function(mat){
  mat_rescaled <- apply(mat, 2, function(v){
    v_range <- range(v)
    names(v_range) <- c("minimum", "maximum")
    range_difference <- v_range["maximum"] - v_range["minimum"]
    rescaled <- (v - v_range["minimum"])/range_difference
    return(rescaled)
  })
  return(mat_rescaled)
}
```

# Getting data from GEOquery

```{r}
geodata <- GEOquery::getGEO(GEO = "GSE76275", destdir = "./tempfiles")
```


```{r}
mdata <- geodata %>%
  pluck(1) %>%
  phenoData() %>%
  pData() %>% as_tibble()
```


```{r}
feature_data <- geodata %>%
  pluck(1) %>%
  featureData()
  
```


```{r}
write_csv(mdata, "raw_mdata.csv")
```


# Inspecting and cleaning the metadata

```{r}
mdata %>% 
  glimpse()
```


```{r}
mdata <- mdata %>% 
  select(title, contains("date"), geo_accession, contains(":ch1"))

colnames(mdata)
  
```


```{r}
cnames <- colnames(mdata)
```


```{r}
cnames_processed <- str_split(cnames, pattern = ":") %>% 
  map_chr(~{.x[[1]]}) %>% 
  str_replace_all(" ", "_") %>% 
  str_replace_all("-", "_") %>% 
  str_remove_all("\\(|\\)|,")

cnames_processed
```


```{r}
colnames(mdata) <- cnames_processed
rm(cnames, cnames_processed)
```


```{r}
glimpse(mdata)
```

```{r}
mdata <- mdata %>% 
  mutate(her2 = if_else(!is.na(her2), her2, "Not Available")) %>% 
  mutate(er = factor(er, levels = c("Negative", "Positive")), 
         pr = factor(pr, levels = c("Negative", "Positive")), 
         her2 = factor(her2, levels = c("Negative", "Positive", "Not Available"))) %>% 
  select(geo_accession, everything())
head(mdata) 
```

Reordering the columns in the metadata.

```{r}
mdata <- select(mdata, geo_accession, everything())
head(mdata)
```

# Reading in raw probe intensity data

Celfiles downloaded from GEO and kept the folder "celfiles/".

```{r}
celFiles <- list.celfiles('celfiles/', full.names = TRUE, listGzipped = TRUE)
celFiles %>% head()
```



```{r}
names(celFiles) <- celFiles %>% 
  basename() %>% 
  str_split("\\.") %>% 
  map_chr(~{.x[1]}) %>% 
  str_split("_") %>% 
  map_chr(~{.x[1]}) 

head(celFiles)
```

Rearranging rows of metadata to match order of samples in `celFiles`.

```{r}
mdata <- mdata[match(mdata$geo_accession, names(celFiles)), ]
```


Getting only the relevant variables from the metadata.

```{r}
mdata_subset <- mdata %>%
  select(geo_accession, 
         title, 
         triple_negative_status, 
         tnbc_subtype,
         submission_date,
         er,
         her2,
         pr,
         race,
         set,
         gender, 
         age_years) %>% 
  mutate(across(where(is.character), .fns = factor)) %>% 
  mutate(tnbc_subtype = if_else(is.na(as.character(tnbc_subtype)), "Not Applicable", as.character(tnbc_subtype))) %>% 
  mutate(tnbc_subtype = factor(tnbc_subtype)) %>% 
  as.data.frame()


rownames(mdata_subset) <- as.character(mdata_subset$geo_accession)

head(mdata_subset)

```

```{r}
rawData <- read.celfiles(celFiles, phenoData = AnnotatedDataFrame(mdata_subset))
```


Looking at the dimensions of the raw expression matrix.

```{r}
exprs(rawData) %>% dim()
```

# Plotting some metadata attributes

Looking at the number of samples for triple negative status and for set.

```{r}
mdata_subset %>% 
  count(triple_negative_status, set)
```



```{r}
mdata_subset %>% 
  count(triple_negative_status) %>% 
  mutate(proportion = round(n/sum(n), 3)) %>% 
  ggplot() +
  geom_col(mapping = aes(y = "dummy_group",
                         x = proportion,
                         fill = triple_negative_status)) +
    theme(axis.text.x = element_text(angle = 90, size = 7),
          axis.text.y = element_blank(),
          axis.ticks.y = element_blank(),
          axis.title.y = element_blank(),
          title = element_text(size = 10),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          panel.background = element_blank(),
          axis.line = element_line(colour = "black", size = 0.5),
          legend.direction = "horizontal", legend.position = "top") +
  geom_text(aes(y = 1, 
                x = proportion, 
                label = proportion, 
                group = triple_negative_status), 
            size = 5, 
            position = position_stack(vjust = 0.5), 
            color = "white") +
  scale_fill_npg(name = "Triple Negative Status") +
  guides(fill = guide_legend(reverse = TRUE)) +
  labs(x = "Proportion",
       title = str_wrap("Proportions of TNBC status values for GSE76275", 60))

ggsave(filename = "plots/exploration_plots/GSE76275_tnbc_proportion_barplot.png")
```

```{r}
mdata_subset %>% 
  count(triple_negative_status, set) %>% 
  mutate(proportion = round(n/sum(n), 3)) %>% 
  ggplot() +
  geom_col(mapping = aes(x = triple_negative_status,
                         y = proportion,
                         fill = set)) +
    geom_text(aes(x = triple_negative_status, 
                y = proportion, 
                label = proportion, 
                group = set), 
            size = 3, 
            position = position_stack(vjust = 0.5), 
            color = "white") +
    theme(title = element_text(size = 10),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          panel.background = element_blank(),
          axis.line = element_line(colour = "black", size = 0.5),
          legend.direction = "vertical", legend.position = "right") +
  scale_fill_npg(name = "Sample Set") +
  labs(x = "TNBC Status",
       y = "Proportion",
       title = str_wrap("Proportions of sample set values for GSE76275", 60))

ggsave(filename = "plots/exploration_plots/GSE76275_set_proportion_barplot.png")
```



```{r}
mdata_subset %>% 
  filter(triple_negative_status == "not TN") %>% 
  count(pr) %>% 
  mutate(proportion = round(n/sum(n), 3)) %>% 
  ggplot() +
  geom_col(mapping = aes(y = "dummy_group",
                         x = proportion,
                         fill = pr )) +
    theme(axis.text.x = element_text(angle = 90, size = 7),
          axis.text.y = element_blank(),
          axis.ticks.y = element_blank(),
          axis.title.y = element_blank(),
          title = element_text(size = 10),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          panel.background = element_blank(),
          axis.line = element_line(colour = "black", size = 0.5),
          legend.direction = "horizontal", legend.position = "top") +
  geom_text(aes(y = 1, 
                x = proportion, 
                label = proportion, 
                group = pr ), 
            size = 5, 
            position = position_stack(vjust = 0.5), 
            color = "white") +
  scale_fill_npg(name = "Progesterone Receptor Status") +
  guides(fill = guide_legend(reverse = TRUE)) +
  labs(x = "Proportion",
       title = str_wrap("Proportions of progesterone receptor status values for non-TNBC samples in GSE7627", 60))


ggsave(filename = "plots/exploration_plots/GSE76275_pr_proportion_barplot.png")
```


```{r}
mdata_subset %>% 
  filter(triple_negative_status == "not TN") %>% 
  count(er) %>%  
  mutate(proportion = round(n/sum(n), 3)) %>% 
  ggplot() +
  geom_col(mapping = aes(y = "dummy_group",
                         x = proportion,
                         fill = er)) +
    theme(axis.text.x = element_text(angle = 90, size = 7),
          axis.text.y = element_blank(),
          axis.ticks.y = element_blank(),
          axis.title.y = element_blank(),
          title = element_text(size = 10),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          panel.background = element_blank(),
          axis.line = element_line(colour = "black", size = 0.5),
          legend.direction = "horizontal", legend.position = "top") +
  geom_text(aes(y = 1, 
                x = proportion, 
                label = proportion, 
                group = er), 
            size = 5, 
            position = position_stack(vjust = 0.5), 
            color = "white") +
  scale_fill_npg(name = "Estrogen Receptor Status") +
  guides(fill = guide_legend(reverse = TRUE)) +
  labs(x = "Proportion",
       title = str_wrap("Proportions of estrogen receptor status values for non-TNBC samples in GSE7627", 60))

ggsave(filename = "plots/exploration_plots/GSE76275_er_proportion_barplot.png")
```

```{r}
mdata_subset %>% 
  filter(triple_negative_status == "not TN") %>% 
  count(her2) %>%  
  mutate(proportion = round(n/sum(n), 3)) %>% 
  ggplot() +
  geom_col(mapping = aes(y = "dummy_group",
                         x = proportion,
                         fill = her2)) +
    theme(axis.text.x = element_text(angle = 90, size = 7),
          axis.text.y = element_blank(),
          axis.ticks.y = element_blank(),
          axis.title.y = element_blank(),
          title = element_text(size = 10),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          panel.background = element_blank(),
          axis.line = element_line(colour = "black", size = 0.5),
          legend.direction = "horizontal", legend.position = "top") +
  geom_text(aes(y = 1, 
                x = proportion, 
                label = proportion, 
                group = her2), 
            size = 5, 
            position = position_stack(vjust = 0.5), 
            color = "white") +
  scale_fill_npg(name = "HER2 Amplification Status") +
  guides(fill = guide_legend(reverse = TRUE)) +
  labs(x = "Proportion",
       title = str_wrap("Proportions of HER2 amplification status values for non-TNBC samples in GSE7627", 60))

ggsave(filename = "plots/exploration_plots/GSE76275_her2_proportion_barplot.png")
```

```{r}
mdata_subset %>% 
  ggplot() +
  geom_boxplot(aes(x = triple_negative_status, y = age_years, fill = set)) +
  scale_fill_npg(name = "Sample Set") +
  theme_light() +
    labs(x = "Triple Negative Status",
         y = "Age in Years",
         title = str_wrap("Age distribution of sample sets for GSE76275", 60))
  

ggsave("plots/exploration_plots/GSE76275_age_boxplot.png")
```



```{r}
list("ER" = mdata_subset$geo_accession[mdata_subset$er == "Positive"],
     "PR" = mdata_subset$geo_accession[mdata_subset$pr == "Positive"],
     "HER2" = mdata_subset$geo_accession[mdata_subset$her2 == "Positive"]) %>%
  fromList(.) %>% 
  upset(., c("ER", "PR", "HER2"), width_ratio = 0.2) +
  ggtitle(str_wrap("Upset plot for different combinations of ER, PR, and HER2 status in the non-TNBC samples in GSE76275", 40)) +
  theme(title = element_text(size = 10))

ggsave("plots/exploration_plots/GSE76275_nonTNBC_upset.png")
```

# Performing RMA

## Using regular RMA on data (without separating by class)

```{r}
res_1 <- rma(rawData)
```

Looking at the dimensions of the expression matrix after RMA.

```{r}
exprs(res_1) %>% 
  dim()
```

Looking at the first 5 rows and columns.

```{r}
exprs(res_1)[1:5, 1:5]
```


## Performing class-specific RMA by reading in the expression sets separately


Getting lists of the TNBC samples and the non-TNBC samples.

```{r}
tnbc_samples <- mdata_subset %>% 
  filter(triple_negative_status == "TN") %>% 
  select(geo_accession) %>% 
  unlist(use.names = F) %>% 
  as.character()

head(tnbc_samples)

nontnbc_samples <- mdata_subset %>% 
  filter(triple_negative_status == "not TN") %>% 
  select(geo_accession) %>% 
  unlist(use.names = F) %>% 
  as.character()

head(nontnbc_samples)
```


Creating different metadata tables for TNBC and nonTNBC.

```{r}
mdata_subset_tnbc <- mdata_subset[tnbc_samples, ]
dim(mdata_subset_tnbc)
mdata_subset_nontnbc <- mdata_subset[nontnbc_samples, ]
dim(mdata_subset_nontnbc)
```


Reading in the TNBC files.

```{r}
rawData_tnbc <- read.celfiles(filenames = celFiles[tnbc_samples],
                              phenoData = AnnotatedDataFrame(mdata_subset_tnbc))

rawData_tnbc
```


Reading in the nonTNBC files.

```{r}
rawData_nontnbc <- read.celfiles(filenames = celFiles[nontnbc_samples],
                              phenoData = AnnotatedDataFrame(mdata_subset_nontnbc))

rawData_nontnbc
```



```{r}
rawData_nontnbc
```


Performing RMA on TNBC data.

```{r}
res_tnbc <- rma(rawData_tnbc)
```


Performing RMA on nonTNBC data.

```{r}
res_nontnbc <- rma(rawData_nontnbc)
```


Combining the expression matrices of TNBC and nonTNBC data after separate RMA.

```{r}
res_joint <- cbind(exprs(res_tnbc), exprs(res_nontnbc))
```

Looking at the first 5 rows and columns of the combined expression matrices.

```{r}
res_joint[1:5, 1:5]
```

# Saving certain CSV files for everyone else to refer to

Saving the joint expression matrix from class-specific QN.

```{r}
res_joint %>% 
  as_tibble(rownames = "probe_id") %>% 
  write_csv("dataframe_files/post_classQN_expression.csv")
```

Saving a subset of the metadata that I think is relevant.

```{r}
mdata_subset %>% 
  write_csv("dataframe_files/metadata_subset.csv")
```


Saving the TNBC and nonTNBC metadata separately, just in case.

```{r}
mdata_subset_tnbc %>% 
  write_csv("dataframe_files/metadata_subset_tnbc.csv")
```


```{r}
mdata_subset_nontnbc %>% 
  write_csv("dataframe_files/metadata_subset_nontnbc.csv")
```


# Getting sample-specific boxplots

## Sample specific boxplots for regular RMA QN


```{r}
res_1_df_long <- res_1 %>%
  exprs() %>% 
  as_tibble(rownames = "probeID") %>% 
  pivot_longer(cols = all_of(c(tnbc_samples, nontnbc_samples)), names_to = "sample_id", 
               values_to = "intensity") %>% 
  left_join(., mdata_subset, by = c("sample_id" = "geo_accession"))
```



```{r}
p2 <- res_1_df_long %>% 
  ggplot() +
  geom_boxplot(mapping = aes(x = reorder(sample_id, as.numeric(set)), y = intensity, color = set), outlier.size = 0.2)
```


```{r}
p2 <- p2 + labs(x = "samples", 
       title = str_wrap("Sample-wise log2 intensity boxplots for global quantile normalization for GSE76275", 60)) +
  scale_color_npg(name = "Triple Negative Status") +
  theme_light() +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        title = element_text(size = 10),
        legend.position = "top", 
        legend.direction = "horizontal")
p2
  
```

```{r}
ggsave("plots/exploration_plots/GSE76275_post_regQN_boxplots.png", 
       p2, 
       units = "cm", 
       width = 30, 
       height = 10)
```


```{r}
rm(res_1_df_long)
```


## Sample specific boxplots for classQN


```{r}
res_joint_df_long <- res_joint %>% 
  as_tibble(rownames = "probeID") %>% 
  pivot_longer(cols = all_of(c(tnbc_samples, nontnbc_samples)), names_to = "sample_id", 
               values_to = "intensity")
  
```


```{r}
saveRDS(object = res_joint_df_long, "res_joint_df_long.RDS")
# res_joint_df_long <- readRDS("res_joint_df_long.RDS")
```



```{r}
p1 <- res_joint_df_long %>% 
  left_join(., mdata_subset, by = c("sample_id" = "geo_accession")) %>% 
  mutate(sample_id = factor(sample_id)) %>% 
  ggplot() +
   geom_boxplot(mapping = aes(x = reorder(sample_id, as.numeric(set)), y = intensity, color = set), outlier.size = 0.2)
```


```{r}
p1 <- p1 + labs(x = "samples", 
       title = str_wrap("Sample-wise log2 intensity boxplots for class-specific quantile normalization for GSE76275", 60)) +
scale_color_npg(name = "Triple Negative Status") +
  theme_light() +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        title = element_text(size = 10),
        legend.position = "top", 
        legend.direction = "horizontal")

p1 
```



```{r}
ggsave("plots/exploration_plots/GSE76275_post_classQN_boxplots.png", 
       p1, 
       units = "cm", width = 20, height = 10)
```



# Performing PCA

## Custom functions

Function to create an annotated data frame by combining PC scores as well as metadata: useful for ggplot visualization.

```{r}
get_pca_annot_df <- function(pca.obj, sample_id_col, mdata_df){
  ind_scores <- pca.obj$x
  ind_scores_reordered <- ind_scores[match(rownames(ind_scores), mdata_df[[sample_id_col]]), ] %>% 
    as_tibble(rownames = sample_id_col) %>% 
    mutate(filename = factor(!!sym(sample_id_col)))
  ind_scores_annot <- left_join(ind_scores_reordered, y = mdata_df, by = sample_id_col) %>% 
  select(all_of(colnames(mdata_subset)), contains("PC"))
  return(ind_scores_annot)
}
```


## Performing PCA on regular RMA data

```{r}
pca.res_1 <- res_1 %>%
  exprs() %>%
  t() %>%
  prcomp(center = TRUE, scale = TRUE)
```


Getting the annotated data frame for the PCA.

```{r}
pca.res_1.annot_df <- get_pca_annot_df(pca.obj = pca.res_1, sample_id_col = "geo_accession", mdata_df= mdata_subset)
head(pca.res_1.annot_df)
```

### Visualizing PCA results

Looking at the variance explained by the first 10 PCs.

```{r}
fviz_eig(pca.res_1) +
  labs(x = "Principal Component", 
       title = str_wrap("Scree plot for the first 10 principal components for global RMA-normalized data for GSE76275", 60)) +
    theme(title = element_text(size = 10), 
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.line = element_line(colour = "black", size = 0.5))

ggsave("plots/exploration_plots/PCA_wholeQN_scree.png", bg = "white")
```

Superimposing variables in data upon sample PCA scores.

The PCA does not seem to separate the TNBC and nonTNBC samples that well when regular RMA is performed.

```{r}
ggplot(pca.res_1.annot_df) + 
  geom_point(mapping = aes(x = PC1, y = PC2, colour = triple_negative_status)) +
    ggtitle(str_wrap("Samples in first two PCs, coloured by triple negative status for GSE76275 after global quantile normalization", 60)) +
  scale_color_npg(name = "Triple Negative Status") +
  guides(colour = guide_legend(override.aes = list(size= 4))) +
  theme(axis.text.x = element_text(angle = 90, size = 7),
               title = element_text(size = 10),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.line = element_line(colour = "black", size = 0.5),
        legend.background = )


ggsave("plots/exploration_plots/PCA_wholeQN_TNBC_status.png", bg = "white")
  
```



The samples do not seem to separate well by set either.

```{r}
ggplot(pca.res_1.annot_df) + 
  geom_point(mapping = aes(x = PC1, y = PC2, colour = set)) +
  ggtitle(str_wrap("Samples in first two PCs, coloured by set (discovery or validation) for GSE76275 after global quantile normalization", 60)) +
  scale_color_aaas(name = "Sample Set") +
  guides(colour = guide_legend(override.aes = list(size= 4))) +
  theme(axis.text.x = element_text(angle = 90, size = 7),
               title = element_text(size = 10),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.line = element_line(colour = "black", size = 0.5))

ggsave("plots/exploration_plots/PCA_wholeQN_set.png", bg = "white")
```


```{r}
ggplot(pca.res_1.annot_df) + 
  geom_point(mapping = aes(x = PC1, y = PC2, colour = tnbc_subtype)) +
    ggtitle("Samples in first two PCs, \ncoloured by tnbc_subtype for whole QN") +
  guides(colour = guide_legend(override.aes = list(size= 4))) +
  theme(axis.text.x = element_text(angle = 90, size = 7),
               title = element_text(size = 10),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.line = element_line(colour = "black", size = 0.5), legend.position = "right", legend.direction = "vertical", legend.key.width = unit(x = 0.5, units = "cm")) 
```

## Performing PCA on separately-performed RMA data

```{r}
pca.res_joint <- res_joint %>% 
  t() %>% 
  prcomp(center = TRUE, scale = TRUE)
```


```{r}
# saveRDS(pca.res_joint, "pca_res_joint.RDS")
pca.res_joint <- readRDS("pca_res_joint.RDS")
```


Getting the annotated data frame for the PCA.

```{r}
pca.res_joint.annot_df <- get_pca_annot_df(pca.obj = pca.res_joint, sample_id_col = "geo_accession", mdata_df= mdata_subset)
head(pca.res_joint.annot_df)
```

### Visualizing PCA results

Looking at the variance explained by the first 10 PCs.

```{r}
fviz_eig(pca.res_joint) +
  labs(x = "Principal Component", 
       title = str_wrap("Scree plot for the first 10 principal components for class-specific RMA-normalized data for GSE76275", 60)) +
    theme(title = element_text(size = 10), 
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.line = element_line(colour = "black", size = 0.5))

ggsave("plots/exploration_plots/PCA_classQN_scree.png")
```

Superimposing variables in data upon sample PCA scores.
The PCA **does** separate the TNBC and nonTNBC samples well when class-specific RMA is performed.

```{r}
ggplot(pca.res_joint.annot_df) + 
  geom_point(mapping = aes(x = PC1, y = PC2, colour = triple_negative_status)) +
    ggtitle(str_wrap("Samples in first two PCs, coloured by triple negative status for GSE76275 after class-specific quantile normalization", 60)) +
  scale_color_npg(name = "Triple Negative Status") +
  guides(colour = guide_legend(override.aes = list(size= 4))) +
  theme(axis.text.x = element_text(angle = 90, size = 7),
               title = element_text(size = 10),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.line = element_line(colour = "black", size = 0.5))

ggsave("plots/exploration_plots/PCA_classQN_TNBC_status.png")
```
The validation nonTNBC samples are separated from the discovery TNBC and validation TNBC samples.

```{r}
ggplot(pca.res_joint.annot_df) + 
  geom_point(mapping = aes(x = PC1, y = PC2, colour = set)) +
  ggtitle(str_wrap("Samples in first two PCs, coloured by set (discovery or validation) for GSE76275 after class-specific quantile normalization", 60)) +
  scale_color_aaas(name = "Sample Set") +
  guides(colour = guide_legend(override.aes = list(size= 4))) +
  theme(axis.text.x = element_text(angle = 90, size = 7),
        title = element_text(size = 10),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.line = element_line(colour = "black", size = 0.5))

ggsave("plots/exploration_plots/PCA_classQN_TNBC_set.png")
```

Submission date is perfectly confounded with TNBC status. May or may not be batch effects.

```{r}
ggplot(pca.res_joint.annot_df) + 
  geom_point(mapping = aes(x = PC1, y = PC2, colour = submission_date)) +
    ggtitle(str_wrap("Samples in first two PCs, coloured by submission date for GSE76275 after class-specific quantile normalization", 60)) +
  scale_color_npg(name = "Submission Date") +
  guides(colour = guide_legend(override.aes = list(size= 4))) +
  theme(axis.text.x = element_text(angle = 90, size = 7),
               title = element_text(size = 10),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.line = element_line(colour = "black", size = 0.5))

ggsave("plots/exploration_plots/PCA_classQN_TNBC_date.png")
```



```{r}
ggplot(pca.res_joint.annot_df) + 
  geom_point(mapping = aes(x = PC1, y = PC2, colour = tnbc_subtype)) +
    ggtitle(str_wrap("Samples in first two PCs, coloured by subtype of TNBC for GSE76275 after class-specific quantile normalization", 60)) +
   scale_color_nejm(name = "TNBC Subtype") +
  guides(colour = guide_legend(override.aes = list(size= 4))) +
  theme(axis.text.x = element_text(angle = 90, size = 7),
               title = element_text(size = 10),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.line = element_line(colour = "black", size = 0.5),
        legend.text = element_text(size = 5), 
        legend.key.height = unit(x = 0.3, units = "cm"),
        legend.key.width = unit(x = 0.3, units = "cm")) +
  guides(color = guide_legend(override.aes = list(size = 1)))

ggsave("plots/exploration_plots/PCA_classQN_TNBC_subtype.png")
```

```{r}
ggplot(pca.res_joint.annot_df) + 
  geom_point(mapping = aes(x = PC1, y = PC2,
                           colour = factor(pr), shape = triple_negative_status)) +
    ggtitle(str_wrap("Samples in first two PCs, coloured by progesterone receptor status for GSE76275 after class-specific quantile normalization", 60)) +
  scale_color_npg(name = "Progesterone Receptor Status") +
  scale_shape_manual(name = "Triple Negative Status", values = c("TN" = 17, "not TN" = 1)) +
  guides(colour = guide_legend(override.aes = list(size= 4))) +
  theme(axis.text.x = element_text(angle = 90, size = 7),
               title = element_text(size = 10),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.line = element_line(colour = "black", size = 0.5),
        legend.title = element_text(size = 7),
        legend.text = element_text(size = 5), 
        legend.key.height = unit(x = 0.3, units = "cm"),
        legend.key.width = unit(x = 0.3, units = "cm")) +
    guides(color = guide_legend(override.aes = list(size = 1)))


ggsave("plots/exploration_plots/PCA_classQN_pr_GSE76275.png", bg= "white")
```

```{r}
ggplot(pca.res_joint.annot_df) + 
  geom_point(mapping = aes(x = PC1, y = PC2,
                           colour = factor(her2), shape = triple_negative_status)) +
    ggtitle(str_wrap("Samples in first two PCs, coloured by HER2 amplification status for GSE76275 after class-specific quantile normalization", 60)) +
  scale_color_nejm(name = "HER2 Amplification Status") +
  scale_shape_manual(name = "Triple Negative Status", values = c("TN" = 17, "not TN" = 1)) +
  theme(axis.text.x = element_text(angle = 90, size = 7),
               title = element_text(size = 10),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.line = element_line(colour = "black", size = 0.5),
        legend.title = element_text(size = 7),
        legend.text = element_text(size = 5), 
        legend.key.height = unit(x = 0.3, units = "cm"),
        legend.key.width = unit(x = 0.3, units = "cm")) + 
      guides(color = guide_legend(override.aes = list(size = 1)))

ggsave("plots/exploration_plots/PCA_classQN_her2_GSE76275.png", bg = "white")
```

```{r}
ggplot(pca.res_joint.annot_df) + 
  geom_point(mapping = aes(x = PC1, y = PC2,
                           colour = factor(er), shape = triple_negative_status)) +
    ggtitle(str_wrap("Samples in first two PCs, coloured by estrogen receptor status for GSE76275 after class-specific quantile normalization", 60)) +
  scale_color_npg(name = "Estrogen Receptor Status") +
  scale_shape_manual(name = "Triple Negative Status", values = c("TN" = 17, "not TN" = 1)) +
  guides(colour = guide_legend(override.aes = list(size= 4))) +
  theme(axis.text.x = element_text(angle = 90, size = 7),
               title = element_text(size = 10),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.line = element_line(colour = "black", size = 0.5),
        legend.title = element_text(size = 7),
        legend.text = element_text(size = 5), 
        legend.key.height = unit(x = 0.3, units = "cm"),
        legend.key.width = unit(x = 0.3, units = "cm")) + 
      guides(color = guide_legend(override.aes = list(size = 1)))


ggsave("plots/exploration_plots/PCA_classQN_er_GSE76275.png")
```

# Performing hierarchical clustering

## Getting distances

```{r}
perform_min_max <- function(x){
  mm_transformation <- preProcess(x, method = "range")
  rescaled <- predict(mm_transformation, x)
  return(rescaled)
}
```


Getting distances after performing min max normalization.


```{r}
# res_1_dists <- exprs(res_1) %>% 
#   t() %>% 
#   perform_min_max() %>% 
#   dist(method = "euclidean")
  
```


```{r}
# saveRDS(res_1_dists, "res_1_dists.RDS")
res_1_dists <- readRDS("res_1_dists.RDS")
```



```{r}
# res_joint_dists <- res_joint %>% 
#     t() %>% 
#   perform_min_max() %>% 
#   dist(method = "euclidean")
  
```


```{r}
# saveRDS(res_joint_dists, "res_joint_dists.RDS")
res_joint_dists <- readRDS("res_joint_dists.RDS")
```



## Using dendrograms

```{r}
res_1_dend <- res_1_dists %>% 
  hclust() %>% 
  as.dendrogram()
```


```{r}
res_joint_dend <- res_joint_dists %>%
  hclust() %>% 
  as.dendrogram()
```

 
```{r}
library(dendextend)
```


```{r}
# res_1_dend %>% 
#   labels()
```


```{r}
# res_1_dend %>% 
#   order.dendrogram()
```

```{r}
# (res_1 %>% 
#   exprs() %>% 
#   colnames())[28]
  
```


```{r}
res_1_dend_laborder <- res_1_dend %>% 
  labels()

```


```{r}
mycolors <- ifelse(mdata_subset[res_1_dend_laborder, ]$triple_negative_status == "TN", "forestgreen", "maroon")
```



```{r}
par(mar = c(10,2,1,1))
res_1_dend %>% 
  set("labels_cex", 0.1) %>% 
  plot()

colored_bars(colors = mycolors, dend = res_1_dend, rowLabels = "TN Status", add = TRUE)
```


```{r}
res_joint_dend_laborder <- res_joint_dend %>% 
  labels()
```


```{r}
mycolors <- ifelse(mdata_subset[res_joint_dend_laborder, ]$triple_negative_status == "TN", "forestgreen", "maroon")
```



```{r}
par(mar = c(10,2,1,1))
res_joint_dend %>% 
  set("labels_cex", 0.1) %>% 
  plot()

colored_bars(colors = mycolors, dend = res_joint_dend, rowLabels = "TN Status", add = TRUE)
```


## Using heatmaps

Function to process distance object into a distance matrix for heatmap visualization.

```{r}
get_distmat <- function(x){
  distmat <- as.matrix(x)
  colnames(distmat) <- NULL
  diag(distmat) <- NA
  return(distmat)
}
```



```{r}
row_annot <- mdata_subset %>% 
  mutate(er = factor(er), pr = factor(pr), her2 = factor(her2)) %>% 
  select(set, submission_date, pr, er, her2) %>% 
    rename(`Sample set` = set, 
           `Submission date` = submission_date,
           `HER2 status` = her2,
           `ER status` = er,
           `PR status` = pr)

head(row_annot)
```


```{r}
set.seed(5)
row_colours <- list( "Sample set" = c("darkgoldenrod4", "darkmagenta", "salmon"), 
                     "Submission date" = pal_nejm()(2),
                     "PR status" = pal_lancet()(9)[1:2], 
                     "ER status" = pal_lancet()(9)[8:9],
                     "HER2 status" = pal_npg()(9)[c(3, 2, 8)])

names(row_colours[["Sample set"]]) <- as.character(unique(row_annot[["Sample set"]]))
names(row_colours[["PR status"]]) <- unique(row_annot[["PR status"]])
names(row_colours[["ER status"]]) <- unique(row_annot[["ER status"]])
names(row_colours[["HER2 status"]]) <- unique(row_annot[["HER2 status"]])
names(row_colours[["Submission date"]]) <- as.character(unique(row_annot[["Submission date"]]))
str(row_colours)
```
```{r}
my_colours <-  viridis(265^2, begin = 1, end = 0)
```

```{r}
res_joint_dists %>% 
  get_distmat() %>% 
pheatmap(.,
         color = my_colours,
         annotation_row = row_annot,
         annotation_colors = row_colours,
         show_colnames = F,
         show_rownames = F,
         cutree_rows = 2,
         cutree_cols = 2,
         main = str_wrap("Heatmap of sample distances for class-specific QN expression matrix for GSE76275", 60),
         legend_labels = c("small distance", "large distance"),
         legend_breaks = c(min(., na.rm = TRUE), 
                         max(., na.rm = TRUE)), 
          filename = "plots/exploration_plots/classQN_clustering_heatmap_GSE76275.png")
```


# Performing SVA

## Performing SVA on regular RMA data


```{r}
full_mod <- mdata_subset %>% 
  select(geo_accession, triple_negative_status) %>% 
  arrange(triple_negative_status) %>% 
  model.matrix(~triple_negative_status, data = .)

head(full_mod)
```


```{r}
red_mod <- model.matrix(~1, data = mdata_subset)

head(red_mod)
```

Get number of significant surrogate variables.

```{r}
n.sv.wholeQN <- num.sv(exprs(res_1), full_mod, method="leek")
```


```{r}
n.sv.wholeQN
```


```{r}
svobj.wholeQN <- sva(exprs(res_1), mod = full_mod, mod0 = red_mod, n.sv = 1)
```


```{r}
sv_df.wholeQN <- tibble("geo_accession" = colnames(exprs(res_1)), "sv" = svobj.wholeQN$sv)

head(sv_df.wholeQN)
```



```{r}
left_join(sv_df.wholeQN, mdata, by = "geo_accession") %>% 
  mutate(index = 5) %>% 
  ggplot() +
  # geom_col(mapping = aes(y = fct_reorder(geo_accession, sv, .fun = function(x){x}), x = sv, fill = set)) +
  geom_boxplot(mapping = aes(x = submission_date, y = sv, fill = set)) +
  theme_light() +
  labs(y = "Surrogate Variable Value", title = "Distribution of latent variable estimated by SVA for different grouping factors")
  

# ggsave("plots/exploration_plots/sva_grouping_normalRMA.png")
```


## Performing SVA on class-specific quantile normalized data

Create full model matrix.

```{r}
full_mod <- mdata_subset %>% 
  select(geo_accession, triple_negative_status) %>% 
  arrange(triple_negative_status) %>% 
  model.matrix(~triple_negative_status, data = .)

head(full_mod)
```

Create reduced model matrix.

```{r}
red_mod <- model.matrix(~1, data = mdata_subset)

head(red_mod)
```

Get number of significant surrogate variables.

```{r}
n.sv.classQN <- num.sv(res_joint, full_mod, method="leek")
```


```{r}
n.sv.classQN
```


Perform SVA on classQN-normalized expression matrix.

```{r}
svobj.classQN <- sva(res_joint, mod = full_mod, mod0 = red_mod, n.sv = n.sv.classQN)
```


```{r}
sv_df.classQN <- tibble("geo_accession" = colnames(res_joint), "sv" = svobj.classQN$sv)

head(sv_df.classQN)
```

```{r}
saveRDS(sv_df.classQN, "sv_df_classQN.RDS")
# sv_df.classQN <- readRDS("sv_df_classQN.RDS")
```


```{r}
sv_df.classQN <- left_join(sv_df.classQN , mdata, by = "geo_accession")
```


```{r}
sv_df.classQN %>% 
ggplot() +
  geom_boxplot(mapping = aes(x = submission_date, y = sv, fill = er)) +
  # geom_point(mapping = aes(x = submission_date, y = sv, color = er)) +
  theme_light() 
  # labs(y = "Surrogate Variable Value", title = "Distribution of latent variable estimated by SVA for different grouping factors")
  

# ggsave("plots/exploration_plots/sva_grouping_classQN.png")
```


# Trying to see if the SVA estimates a batch when QN is not applied

In this attempt, I perform no quantile normalization while performing RMA. If QN has not been performed and a surrogate variable shows up that corresponds to batch, batch effects are probably present.

```{r}
rawData.summary <- rma(rawData, background = TRUE, normalize = FALSE)
```

```{r}
# saveRDS(rawData.summary, "rawData_summary.RDS")
rawData.summary <- readRDS("rawData_summary.RDS")
```


```{r}
rawData.summary_df_long <- rawData.summary %>% 
  exprs() %>% 
  as_tibble(rownames = "probeID") %>% 
  pivot_longer(cols = all_of(c(tnbc_samples, nontnbc_samples)), names_to = "sample_id", 
               values_to = "intensity") %>% 
  left_join(., mdata_subset, by = c("sample_id" = "geo_accession"))
```


```{r}
# saveRDS(rawData.summary_df_long, "rawData_summary_df_long.RDS")
rawData.summary_df_long <- readRDS("rawData_summary_df_long.RDS")
```


```{r}
p3 <- rawData.summary_df_long %>% 
  ggplot() +
  geom_boxplot(mapping = aes(x = reorder(sample_id, as.numeric(set)), y = intensity, 
                             color = set))
```


```{r}
p3 <- p3 + labs(x = "samples", 
       title = str_wrap("Sample-wise log2 intensity boxplots for non- quantile normalized expression values for GSE76275", 60)) +
  scale_color_npg(name = "Triple Negative Status") +
  theme_light() +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        title = element_text(size = 10),
        legend.position = "top", 
        legend.direction = "horizontal")

p3
```


```{r}
ggsave("plots/exploration_plots/GSE76275_noQN_boxplots.png", 
       p3, 
       units = "cm", width = 30, height = 10)
```


Create full model matrix.

```{r}
full_mod <- mdata_subset %>% 
  select(geo_accession, triple_negative_status) %>% 
  arrange(triple_negative_status) %>% 
  model.matrix(~triple_negative_status, data = .)

head(full_mod)
```

Create reduced model matrix.

```{r}
red_mod <- model.matrix(~1, data = mdata_subset)

head(red_mod)
```

Getting the number of surrogate variables in the absence of quantile normalization.

```{r}
n.sv.nonorm <- num.sv(exprs(rawData.summary), full_mod, method="leek")
```

There is one surrogate variable present in the absence of QN.

```{r}
n.sv.nonorm
```



```{r}
svobj.nonorm <- sva(exprs(rawData.summary), mod = full_mod, mod0 = red_mod, n.sv = n.sv.nonorm)
```


```{r}
sv_df.nonorm <- tibble("geo_accession" = colnames(exprs(rawData.summary)), "sv" = svobj.nonorm$sv)

head(sv_df.nonorm)
```

```{r}
saveRDS(sv_df.nonorm, "sv_df_nonorm.RDS")
# sv_df.nonorm <- readRDS("sv_df_nonorm.RDS")
```


```{r}
left_join(sv_df.nonorm, mdata_subset, by = "geo_accession") %>% 
  mutate(index = 5) %>% 
  ggplot() +
  # geom_col(mapping = aes(y = fct_reorder(geo_accession, sv, .fun = function(x){x}), x = sv, fill = set)) +
  geom_boxplot(mapping = aes(x = submission_date, y = sv, fill = set)) +
  scale_fill_npg(name = "Sample Set") +
  theme_light() +
  labs(x = "Submission Date", 
         y = "Surrogate Variable Value", 
       title = str_wrap("Distribution of latent variable estimated by SVA for submission dates, coloured by sample set", 50))

ggsave("plots/exploration_plots/sva_grouping_date_noQN.png")
```


```{r}
left_join(sv_df.nonorm, mdata_subset, by = "geo_accession") %>% 
  mutate(index = 5) %>% 
  ggplot() +
  geom_boxplot(mapping = aes(x = pr, y = sv, fill = set)) +
  scale_fill_npg(name = "Sample Set") +
  theme_light() +
   labs(x = "Progesterone Receptor Status",
    y = "Surrogate Variable Value",
       title = str_wrap("Distribution of latent variable estimated by SVA for progesterone receptor status, coloured by sample set", 50))
 

ggsave("plots/exploration_plots/sva_grouping_pr_noQN.png")
```


```{r}
left_join(sv_df.nonorm, mdata_subset, by = "geo_accession") %>% 
  mutate(index = 5) %>% 
  ggplot() +
  geom_boxplot(mapping = aes(x = er, y = sv, fill = set)) +
  scale_fill_npg(name = "Sample Set") +
  theme_light() +
   labs(x = "Estrogen Receptor Status",
    y = "Surrogate Variable Value",
       title = str_wrap("Distribution of latent variable estimated by SVA for estrogen receptor status, coloured by sample set", 50))
 

ggsave("plots/exploration_plots/sva_grouping_er_noQN.png")
```

```{r}
left_join(sv_df.nonorm, mdata_subset, by = "geo_accession") %>% 
  mutate(index = 5) %>% 
  ggplot() +
  geom_boxplot(mapping = aes(x = her2, y = sv, fill = set)) +
  scale_fill_npg(name = "Sample Set") +
  theme_light() +
   labs(x = "HER2 Amplification Status",
    y = "Surrogate Variable Value",
       title = str_wrap("Distribution of latent variable estimated by SVA for HER2 amplification status, coloured by sample set", 50))
 

ggsave("plots/exploration_plots/sva_grouping_her2_noQN.png")
```



```{r}
left_join(sv_df.nonorm, mdata_subset, by = "geo_accession") %>% 
  mutate(index = 5) %>% 
  ggplot() +
  geom_point(mapping = aes(x = as.numeric(age_years), y = sv, color = set)) +
  scale_color_npg(name = "Sample set") +
  theme(axis.text.x = element_text(angle = 90, size = 7),
        title = element_text(size = 10),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.line = element_line(colour = "black", size = 0.5),
        legend.title = element_text(size = 7),
        legend.text = element_text(size = 5), 
        legend.key.height = unit(x = 0.3, units = "cm"),
        legend.key.width = unit(x = 0.3, units = "cm")) + 
      guides(color = guide_legend(override.aes = list(size = 1))) +
  labs(x = "Age in Years",
       y = "Surrogate Variable Value",
       title = str_wrap("Scatterplot of age latent variable estimated by SVA versus age", 50))

ggsave("plots/exploration_plots/sva_grouping_age_noQN.png")
```

