---
title: "Performing GFS on , and getting DEGs"
output: 
  html_notebook:
    toc: true
    toc_depth: 2
    toc_float: true
---

# Loading libraries


```{r}
library(GEOquery)
library(oligo)
library(tidyverse)
library(ggsci)
# library(pheatmap)
library(RColorBrewer)
# library(viridis)
```

# Custom functions

```{r}
# given a matrix, perform min-max scaling on its columns
min_max_mat <- function(mat){
  mat_rescaled <- apply(mat, 2, function(v){
    v_range <- range(v)
    names(v_range) <- c("minimum", "maximum")
    range_difference <- v_range["maximum"] - v_range["minimum"]
    rescaled <- (v - v_range["minimum"])/range_difference
    return(rescaled)
  })
  return(mat_rescaled)
}
```

## GFS functions

```{r}
# Given a vector and the theta 1 and theta 2 values, return a vector with the fuzzy score
fuzzy_score_ranks <- function(x, theta_1, theta_2){
  ranked_x <- rank(desc(x))
  q1 <- quantile(ranked_x, theta_1)
  q2 <- quantile(ranked_x, theta_2)

  q_diff <- q1 - q2
  scored <- map_dbl(ranked_x, ~{
    if(.x < q1){
      return(1)
    } else if(.x > q1 && .x <= q2){
      fuzzy_score <- (.x - q2)/q_diff
      return(fuzzy_score)
    } else {
      return(0)
    }
  })
  return(scored)
}
```




```{r}
# given a matrix and theta 1 and theta 2, return a fuzzy-scored matrix
## Uses the fuzzy_score_ranks function defined above on every column vector of the matrix
perform_gfs <- function(emat, theta_1, theta_2){
  sample_ids <- colnames(emat)
  probe_ids <- rownames(emat)
  # iterating over the columns of the matrix to return a list of vectors
  scored_list <- map(1:ncol(emat), .f = ~{
    col_values <- emat[, .x]
    ranked_col <- rank(col_values)
    scored_column <- fuzzy_score_ranks(ranked_col, theta_1, theta_2)
    return(scored_column)
  }, emat = emat)
  # column bind the list of vectors to create a matrix
  scored_mat <- do.call(cbind, scored_list)
  # return the annotation to the new fuzzy-scored matrix
  colnames(scored_mat) <- sample_ids
  rownames(scored_mat) <- probe_ids
  return(scored_mat)
}
```

Create a test matrix.

```{r}
temp_emat <- replicate(n = 10, expr = sample(c(100:110), 10, replace = FALSE), simplify = TRUE)
emat
```
Annotate the test matrix.

```{r}
sample_ids <- paste("sample", 1:10, sep = "_")
colnames(temp_emat) <- sample_ids
rownames(temp_emat) <- letters[1:10]
temp_emat
```


```{r}
perform_gfs(temp_emat, 0.10, 0.20)
```

# Prepare the metadata

## Getting data from GEOquery

```{r}
# geodata <- GEOquery::getGEO(GEO = "GSE76275", destdir = "./tempfiles")
# geodata <- GEOquery::getGEO(filename = "./tempfiles/GSE76275_series_matrix.txt.gz")
```


```{r}
# saveRDS(geodata, "geodata.RDS")
geodata <- readRDS("geodata.RDS")
```


```{r}
# mdata <- geodata %>% 
#   pluck(1) %>% 
#   phenoData() %>%
#   pData() %>% as_tibble()
```


```{r}
# feature_data <- geodata %>% 
#   pluck(1) %>% 
#   featureData()
  
```


```{r}
# write_csv(mdata, "raw_mdata.csv")
mdata <- read_csv("raw_mdata.csv")
```


```{r}
# saveRDS(feature_data, "featureData.RDS")
# feature_data <- readRDS("featureData.RDS")
```

## Cleaning the metadata


```{r}
mdata <- mdata %>% 
  select(title, contains("date"), geo_accession, contains(":ch1"))

cnames <- colnames(mdata)

cnames_processed <- str_split(cnames, pattern = ":") %>% 
  map_chr(~{.x[[1]]}) %>% 
  str_replace_all(" ", "_") %>% 
  str_replace_all("-", "_") %>% 
  str_remove_all("\\(|\\)|,")

cnames_processed

colnames(mdata) <- cnames_processed
rm(cnames, cnames_processed)
  
```

```{r}
glimpse(mdata)
```


# Reading in raw probe intensity data

Celfiles downloaded from GEO and kept the folder "celfiles/"

```{r}
celFiles <- list.celfiles('celfiles/', full.names = TRUE, listGzipped = TRUE)
celFiles %>% head()
```



```{r}
names(celFiles) <- celFiles %>% 
  basename() %>% 
  str_split("\\.") %>% 
  map_chr(~{.x[1]}) %>% 
  str_split("_") %>% 
  map_chr(~{.x[1]}) 

head(celFiles)
```
Rearranging rows of metadata to fit the order of the celfiles.

```{r}
mdata <- mdata[match(mdata$geo_accession, names(celFiles)), ]
```


Getting only the relevant variables from the metadata.

```{r}
mdata_subset <- mdata %>%
  select(geo_accession, 
         title, 
         triple_negative_status, 
         tnbc_subtype,
         submission_date,
         er,
         her2,
         pr,
         race,
         set,
         gender, 
         age_years) %>% 
  mutate(across(where(is.character), .fns = factor)) %>% 
  mutate(tnbc_subtype = if_else(is.na(as.character(tnbc_subtype)), "Not Applicable", as.character(tnbc_subtype))) %>% 
  mutate(tnbc_subtype = factor(tnbc_subtype)) %>% 
  as.data.frame()


rownames(mdata_subset) <- as.character(mdata_subset$geo_accession)

head(mdata_subset)

```

```{r}
# rawData <- read.celfiles(celFiles, phenoData = AnnotatedDataFrame(mdata_subset))
```


```{r}
# saveRDS(object = rawData, "rawData.RDS")
rawData <- readRDS("rawData.RDS")
```

# Getting expression values from raw data

```{r}

```

